Parsing Postfix log files

### 2009-03-02 doc/logparser-architecture.tex
Improved the Framework section of the Rules chapter, and made some minor supporting changes elsewhere.

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix may vary widely from host
to host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the architecture's design aims was to make adding
new rules to parse new inputs as easy as possible, to enable administrators
to properly parse their own log files.  To enable this the architecture is
divided into three parts: Framework, Actions, and Rules.  Each will be
discussed separately, but first an overview:

\begin{eqlist}

@@

    \item [Actions] Each action performs the work required to deal with a
        single category of inputs, e.g.\ rejecting a delivery attempt.
        Actions are invoked to process an input once it has been classified
        by a rule.

    \item [Rules]  The rules are responsible for classifying inputs; each
        rule recognises one input variant, and specifies the action that
        will be invoked to process the input. Rules provide an easily
        extensible method of associating inputs with actions.

@@

Unlike traditional parsers such as those used when compiling a programming
language, this architecture does not require a fixed grammar specification
that inputs must adhere to.  The architecture is capable of dealing with
interleaved inputs, out of order inputs, and ambiguous inputs where
heuristics must be applied --- all have arisen and been successfully
accommodated in the Postfix log file parser.

@@

XXX EXTEND FRAMEWORK SECTION\@: EXPAND ON ALL OF THE POINTS IN THE LIST IN
THE NEXT PARAGRAPH\@.

@@

The framework manages the parsing process and provides support functions,
freeing the programmers writing actions to concentrate on writing
productive code.  It links actions and rules, allowing either to be
improved independently of the other, and allows new rules to be written
without needing changes to the source code of a parser.  The framework is
the core of the architecture and is deliberately quite simple: the rules
deal with the variation in inputs, and the actions deal with the
intricacies and complications encountered during parsing.

The function that finds the rule matching the input and invokes the
requested action can be expressed in pseudo-code as:

\begin{verbatim}
INPUT:
for each input {
    for each rule defined by the user {
        if this rule matches the input {
            perform the action specified by the rule
            next INPUT
        }
    }
    warn the user that the input was not parsed
}
\end{verbatim}

Most parsers will require the same basic functionality from the framework,
plus some specialised support functions.  XXX EXPAND THIS\@.

XXX TURN THIS INTO A LIST AND EXPAND THE ITEMS\@.  It provides shared
storage to pass data between actions, saves and loads state, loads and
validates rules, manages parsing, invokes actions, stores results of
parsing, and miscellaneous other tasks.

@@

@@

Glucose: $C_{6}H_{12}O_{6}$
Hydrocarbon: $CH_{3}{(CH_{2})}_{50}CH_{3}$

@@

    % Reduce inter-item spacing because each item is a single line.
    \setlength{\itemsep}{0pt}

>>>>

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix may vary widely from host
to host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the architecture's design aims was to make adding
new rules to parse new inputs as easy as possible, to enable administrators
to properly parse their own log files.  The solution developed is to divide
the architecture into three parts: Framework, Actions, and Rules.  Each
will be discussed separately, but first an overview:

\begin{eqlist}[\def\makelabel#1{\bfseries#1}]

@@

    \item [Actions] Each action performs the work required to deal with a
        single \textit{category\/} of inputs, e.g.\ rejecting a delivery
        attempt.  Actions are invoked to process an input once it has been
        classified by a rule.

    \item [Rules]  The rules are responsible for classifying inputs; each
        rule recognises one input \textit{variant}, and specifies the
        action that will be invoked to process the input.  Each input
        category may have many input variants.  Rules provide an easily
        extensible method of associating inputs with actions.

@@

Unlike traditional parsers such as those used when compiling a programming
language, this architecture does not require a fixed grammar specification
that inputs must adhere to.  The architecture is capable of dealing with
interleaved inputs, out of order inputs, and ambiguous inputs where
heuristics must be applied --- all have arisen and been successfully
accommodated in the Postfix log file parser.  This architecture is ideally
suited to parsing inputs where the input is not fully understood or does
not conform to a fixed grammar: the architecture warns about unparsed
inputs and other errors, but continues parsing as best it can, allowing the
developer of a new parser to decide which deficiencies are most important
and the order to address them in, rather than being forced to fix the first
error that arises.

@@

@@

The framework manages the parsing process and provides support functions,
freeing the programmers writing actions to concentrate on writing
productive code.  It links actions and rules, allowing either to be
improved independently of the other, and allows new rules to be written
without needing changes to the source code of a parser.  The framework is
the core of the architecture and is deliberately quite simple: the rules
deal with the variation in inputs, and the actions deal with the
intricacies and complications encountered during parsing.  The function
that finds the rule matching the input and invokes the requested action can
be expressed in pseudo-code as:

\begin{verbatim}
INPUT:
for each input {
    for each rule defined by the user {
        if this rule recognises the input {
            perform the action specified by the rule
            next INPUT
        }
    }
    warn the user that the input was not parsed
}
\end{verbatim}

Most parsers will require the same basic functionality from the framework;
it is responsible for managing the parsing process from start to finish,
which generally will involve the following steps:

\begin{description}

    \item [Register actions]  Each action needs to be registered with the
        framework so that the framework knows about it; the list of
        registered actions will alsobe used when validating rules.

    \item [Load and validate rules]  The framework loads the rules from
        wherever they are stored: a simple file, a database, or possibly
        even a web server or other network service --- though that would
        have serious security implications.  It validates each rule to
        hopefully catch problems as early in the parsing process as
        possible; the checks will be implementation-specific to some
        extent, but will generally include some of the following:

        \begin{itemize}

            \squeezeitems{}

            \item Ensuring the specified action has been registered.

            \item Checking for conflict in the data to be extracted, e.g.\
                setting the same variable twice.

            \item Checking that the regex used in the rule is valid.

            \item Some optimisation steps may also be performed during
                loading of rules, as discussed in \sectionref{parser
                efficiency}.

        \end{itemize}

    \item [Classify the input]  The pseudo-code above shows how rules are
        successively tried until one is found that recognises the current
        input.  That pseudo-code is very simple: there may be efficiency
        concerns, (\sectionref{parser efficiency}), rule conditions
        (\sectionref{attaching conditions to rules}), or prioritisations
        (\sectionref{prioritisation and rule ordering}) that complicate the
        process.

    \item [Invoke actions]  Once a rule has been found that recognises the
        current input, the specified action must be invoked.  The framework
        marshals the data extracted by the rule, invokes the action, and
        repeats the parsing process if cascaded parsing
        (\sectionref{cascaded parsing}) is used.

    \item [Shared storage]  Parsers generally need to save some state
        information about the input being parsed, e.g.\ a compiler tracking
        which variables are in lexical scope as it moves from one lexical
        block to another.  The framework provides shared storage to deal
        with this and any other storage needs the actions have.  Actions
        may need to exchange data to correctly parse the input, e.g.\
        setting or clearing flags, maintaining a list of previously used
        identifiers, or ensuring at a higher level that the input being
        parsed meets a list of requirements.

    \item [Save and load state]  The architecture can save current state,
        viz the shared storage it provides for actions, and reload it
        later, so that information is not lost between parsing runs.

    \item [Specialised support functions]  Actions may need support or
        utility functions, and in general those functions should be
        available to all action.  Depending on the implementation the
        framework may be a good location for support functions; if there is
        another way to make those functions available to all actions it may
        be preferable to use it, maintaining a clear separation of
        concerns.

\end{description}

@@

\subsection{Cascaded Parsing}

\label{cascaded parsing}

@@

Glucose: $C_{6}H_{12}O_{6}$
Hydrocarbon (polymer): $CH_{3}{(CH_{2})}_{50}CH_{3}$

@@

    % Reduce inter-item spacing because each item is a single line.
    \squeezeitems{}
